Text embedding models yield high-resolution insights into conceptual knowledge from short multiple-choice quizzes

### 2023-10-22 paper/main.tex

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive
neuroscience (e.g.,~\citealp{Kato40, Gall00, ScotEtal07, HallGree08, Macl05}),
has profound analogs in the fields of natural language processing and natural
language understanding. For example, considering the raw contents of a document
(e.g., its constituent symbols, letters, and words) might provide some
clues as to what the document is about, just as memorizing a passage
might provide some ability to answer simple questions about it. However, text
embedding models (e.g.,~\citealp{LandDuma97, DeerEtal90, BleiEtal03,
BleiLaff06, MikoEtal13a, CerEtal18, BrowEtal20}) also attempt to capture the
deeper meaning \textit{underlying} those atomic elements. These models consider
not only the co-occurrences of those elements within and across documents, but
also patterns in how those elements appear across different scales (e.g.,
sentences, paragraphs, chapters, etc.), the temporal and grammatical properties
of the elements, and other high-level characteristics of how they are
used~\citep{Mann20, Mann21a}. According to these models, the deep conceptual
meaning of a document may be captured by a feature vector in a high-dimensional
representation space, wherein nearby vectors reflect conceptually related
documents. A model that succeeds at capturing an analogue of ``understanding'' is
able to assign nearby feature vectors to two conceptually related documents,
\textit{even when the specific words contained in those documents have very little
overlap}.

Given these insights, what form might a representation of the sum total of a
person's knowledge take? First, we might require a means of systematically
describing or representing the nearly infinite set of possible things a person
could know. Second, we might want to account for potential associations between
different concepts. For example, the concepts of ``fish'' and ``water'' might
be associated in the sense that fish live in water. Third, knowledge may have a
critical dependency structure, such that knowing about a particular concept
might require first knowing about a set of other concepts. For example,
understanding the concept of a fish swimming in water first requires
understanding what fish and water \textit{are}. Fourth, as we learn, our
``current state of knowledge'' should change accordingly. Learning new concepts
should both update our characterizations of ``what is known'' and also
unlock any now-satisfied dependencies of those newly learned concepts so that
they are ``tagged'' as available for future learning.

@@

@@

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). We then hand-created 39 multiple-choice
questions: 15 about the conceptual content of \textit{Four Fundamental Forces}
(i.e., Lecture~1), 15 about the conceptual content of \textit{Birth of Stars}
(i.e., Lecture~2), and 9 questions that tested for general conceptual knowledge
about basic physics (covering material that was not presented in either video).
The full set of questions and answer choices may be found in Supplementary
Table~\questions.

>>>>

The difference between ``understanding'' and ``memorizing,'' as framed by
researchers in education, cognitive psychology, and cognitive neuroscience
(e.g.,~\citealp{Kato40, Gall00, ScotEtal07, HallGree08, Macl05}), has profound
analogs in the fields of natural language processing and natural language
understanding. For example, considering the raw contents of a document (e.g.,
its constituent symbols, letters, and words) might provide some clues as to
what the document is about, just as memorizing a passage might provide some
ability to answer simple questions about it. However, text embedding models
(e.g.,~\citealp{LandDuma97, DeerEtal90, BleiEtal03, BleiLaff06, MikoEtal13a,
CerEtal18, BrowEtal20}) also attempt to capture the deeper meaning
\textit{underlying} those atomic elements. These models consider not only the
co-occurrences of those elements within and across documents, but (in many
cases) also patterns in how those elements appear across different scales
(e.g., sentences, paragraphs, chapters, etc.), the temporal and grammatical
properties of the elements, and other high-level characteristics of how they
are used~\citep{Mann20, Mann21a}. According to these models, the deep
conceptual meaning of a document may be captured by a feature vector in a
high-dimensional representation space, wherein nearby vectors reflect
conceptually related documents. A model that succeeds at capturing an analogue
of ``understanding'' is able to assign nearby feature vectors to two
conceptually related documents, \textit{even when the specific words contained
in those documents have very little overlap}.

Given these insights, what form might a representation of the sum total of a
person's knowledge take (speculatively)? First, we might require a means of
systematically describing or representing the nearly infinite set of possible
things a person could know. Second, we might want to account for potential
associations between different concepts. For example, the concepts of ``fish''
and ``water'' might be associated in the sense that fish live in water. Third,
knowledge may have a critical dependency structure, such that knowing about a
particular concept might require first knowing about a set of other concepts.
For example, understanding the concept of a fish swimming in water first
requires understanding what fish and water \textit{are}. Fourth, as we learn,
our ``current state of knowledge'' should change accordingly. Learning new
concepts should both update our characterizations of ``what is known'' and also
unlock any now-satisfied dependencies of those newly learned concepts so that
they are ``tagged'' as available for future learning.

@@

It is important to clarify that although we use topic model-derived embeddings
to \textit{characterize} the conceptual content of the lectures and questions,
we do not claim that the topic model itself \textit{understands} the conceptual
content of the lectures or questions. Rather, we view the topic model as a tool
for capturing the \textit{structure} of the conceptual content of the lectures
and questions in a way that enables us to capture, quantify, and track and
predict participants' knowledge.

@@

We view our work as making several contributions to the study of how people
acquire conceptual knowledge. First, from a methodological standpoint, our
modeling framework provides a systematic means of mapping out and
characterizing knowledge in maps that have infinite (arbitrarily many) numbers
of coordinates, and of ``filling out'' those maps using relatively small
numbers of multiple choice quiz questions. Our experimental finding that we can
use these maps to predict responses to held-out questions
(Fig.~\ref{fig:predictions}) also has important psychological implications. One
such psychological implication is that concepts that are assigned to nearby
coordinates by the text embedding model also appear to be ``known to a similar
extent'' (as reflected by participants' responses to held-out questions). This
suggests that participants also \textit{conceptualize} similarly the content
reflected by nearby embedding coordinates. A second psychological implication
is that something about how participants' knowledge ``spreads'' across concepts
is being captured by the knowledge maps we are inferring from their quiz
responses (e.g., Fig.~\ref{fig:smoothness}). In other words, our study shows
that knowledge about a given concept implies knowledge about related concepts.

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). We then hand-created 39 multiple-choice
questions: 15 about the conceptual content of \textit{Four Fundamental Forces}
(i.e., Lecture~1), 15 about the conceptual content of \textit{Birth of Stars}
(i.e., Lecture~2), and 9 questions that tested for general conceptual knowledge
about basic physics (covering material that was not presented in either video).
We note that estimating the specific ``amount'' of conceptual understanding
that each question ``requires'' to answer is somewhat subjective, and might
even come down to the ``strategy'' a given participant uses to answer the
question at that particular moment. The full set of questions and answer
choices may be found in Supplementary Table~\questions.
